---
title: "Intro to R Markdown"
author: "Nicholas Pearson"
date: "2025-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
R Markdown allows you to combine narrative text, code, and output in the same document. It supports flexible formatting options and control over code visibility.


### Code Chunks
R code is included in chunks using triple backticks. Backticks (ASCII 96) are not available on the italian keyboard so you can create a new chunk either using the GUI or by pressing `Ctrl+Alt+I` on a Windows machine. You can also name your chunks **but no two chunks can have the same name**.
```{r cars}
summary(cars)
```
In is also possible to run Python code.
```{python}
y = range(0, 10)
for i in y:
  print(i)
```
Chunk visibility:

- `echo = TRUE`: show both code and output
- `echo = FALSE`: hide code but show output
- `include = FALSE`: run code, but show neither code nor output
- `eval = FALSE`: show code, but don’t run it
- `results = 'hide'`: run code but hide output

### Text
#### Text Formatting

- **Bold** `**bold**` $\rightarrow$ bold
- *Italic* `*italic*` $\rightarrow$ italic
- <u>Underline</u> `<u>Underline</u>` $\rightarrow$ underline
- `Code effect` ` `Code Effect` ` $\rightarrow$ code effect

### Lists
**Bullet Points**

- B1
- B2

**Numbered Lists**

1. One
2. Two

### Equations
LaTeX-style math is supported:

- Inline $\sigma = \sqrt{\sigma^2}$
- Display (or centered)

$$Z = \cfrac{X-\mu}{\sigma}$$
A comprehensive cheat-sheet for LaTeX commands can be found here: `https://quickref.me/latex`.

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot. By removing `echo = False` we now get:
```{r}
plot(pressure)
```
