---
title: Data Analysis
author: "Nicholas Pearson"
date: "September 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup
```{r cars}
#Load the required libraries
library(insuranceData)
library(knitr)
```

## Load the data
We will analyze the `AutoClaims` dataset from the insuranceData package, which contains insurance claim information. More details about this dataset can be found in the insuranceData package documentation. 
```{r pressure, echo=FALSE}
# Load the AutoClaims dataset
data(AutoClaims)

# Display first few rows
head(AutoClaims)
```

Here we can see the first few rows of the dataset, giving us an overview of the variables.
## Initial Analysis
Let’s examine the structure and summary statistics of the dataset. This will help us understand the types of variables and their distributions.
```{r}
# Structure of the datset
str(AutoClaims)

# Summary of the variables
summary(AutoClaims)
```

## Frequency Distribution of Risk Classes
Compute the absolute frequency distribution, the relative frequency distribution and the percentage frequency distribution of the variable CLASS. Print the results in three columns.
```{r}
# Absolute frequencies
abs_freq <- table(AutoClaims$CLASS)
abs_freq

# Relative frequencies
rel_freq <- prop.table(abs_freq)
rel_freq

# Percentage Frequencies
perc_freq <- prop.table(abs_freq) * 100
perc_freq
```

A `bar plot` is a suitable representation for categorical variables like CLASS:
```{r}
barplot(abs_freq, 
        main = "Distribution of Insurance Risk Classes",
        xlab = "Risk Class",
        ylab = "Frequency",
        col = "steelblue",
        las = 2)  # Rotate x-axis labels
```

## Distribution of Claim Amounts (`PAID`)

We now look at the PAID variable, representing claim amounts.
We can represent it graphically using a `histogram`.
```{r}
# Basic histogram for PAID variable
hist(AutoClaims$PAID, 
     breaks = 50,
     main = "Distribution of Insurance Claim Costs",
     xlab = "Claim Cost ($)",
     ylab = "Frequency",
     col = "lightblue",
     border = "black")
```

We can add mean, median, and quartiles to better understand the spread:
```{r}
mean(AutoClaims$PAID)
median(AutoClaims$PAID)

hist(AutoClaims$PAID, 
     breaks = 50,
     main = "Distribution of Insurance Claim Costs",
     xlab = "Claim Cost ($)",
     ylab = "Frequency",
     col = "lightblue",
     border = "black",
     xlim = c(0, 10000)
 )
abline(v = mean(AutoClaims$PAID), col = "red", lwd = 2, lty = 2) 
abline(v = median(AutoClaims$PAID), col = "blue", lwd = 2, lty = 2)
abline(v = quantile(AutoClaims$PAID, 0.25), col = "orange", lwd = 2, lty = 2)
abline(v = quantile(AutoClaims$PAID, 0.75), col = "orange", lwd = 2, lty = 2)
```

Red line: mean; Blue line: median; Orange lines: 1st and 3rd quartiles.

## Frequency Distribution in Classes

We create custom intervals for `PAID` to better understand the distribution of claims considering the following classes: [0,5000] (5000,10000] (10000,20000] (20000,50000] (50000,100000] (100000,+)
```{r}
# Define class intervals
breaks <- c(0, 5000, 10000, 20000, 50000, 100000, Inf)
labels <- c("[0,5000]", "(5000,10000]", "(10000,20000]", 
            "(20000,50000]", "(50000,100000]", "(100000,+)")


# Create classes
AutoClaims$PAID_classes <- cut(AutoClaims$PAID, 
                               breaks = breaks, 
                               labels = labels, 
                               include.lowest = TRUE)
AutoClaims$PAID_classes[1:10]

# Compute frequency distributions
abs_freq_paid <- table(AutoClaims$PAID_classes)
rel_freq_paid <- prop.table(abs_freq_paid)
perc_freq_paid <- prop.table(abs_freq_paid) * 100

abs_freq_paid
```

Visualizing these classes:
```{r}
barplot(abs_freq_paid, 
        main = "Distribution of PAID",
        xlab = "Paid",
        ylab = "Frequency",
        col = "steelblue",
        las = 2)  # Rotate x-axis labels
```

Feel free to experiment with other intervals to see how the distribution changes.
We can also compute densities to normalize the distributions by class width:
```{r}
#Compute class widths
class_widths <- diff(breaks)
#Given that the last one has no upper bound set the width of the second to last one
class_widths[length(class_widths)] <- class_widths[length(class_widths)-1]  # Use previous width for last open interval
#Create densities
densities <- as.numeric(rel_freq_paid) / class_widths

barplot(densities,
        names.arg = names(abs_freq_paid),
        main = "Density Plot of Insurance Claim Costs by Classes",
        xlab = "Cost Classes ($)",
        ylab = "Density",
        col = "lightgreen",
        las = 2)
```

## Comparing Claim Amounts by Gender

We can compare claim distributions for males and females:
```{r}
par(mfrow=c(1,2))

hist(AutoClaims$PAID[AutoClaims$GENDER=="M"], col = "lightblue", 
     include.lowest = TRUE, main="MALE",
     ylim = c(0, 4000),
      xlab = "Paid Claim")
hist(AutoClaims$PAID[AutoClaims$GENDER=="F"], col = "lightpink",
     include.lowest = TRUE, main="FEMALE",
     ylim = c(0, 4000),
     xlab = "Paid Claim")

```

## Age and Gender Comparison

`Boxplots` help visualize the age distribution by gender:
```{r}
boxplot(AGE ~ GENDER, data = AutoClaims,
        main = "Age by Gender",
        xlab = "Gender",
        ylab = "Age",
        col = c("lightpink", "lightblue"),
        border = "black",
        notch = FALSE)
```

## Claim Amount by Risk Class

Finally, we compare the square root of `PAID` by `CLASS` to improve readability:
```{r}
boxplot(sqrt(PAID) ~ CLASS, data = AutoClaims,
        main = "Square root of PAID by CLASS",
        xlab = "Class",
        ylab = "sqrt Paid",
        border = "black",
        notch = FALSE)
```